noam_data_exploration_v1

### 2024-07-16 noam_data_exploration_v3.Rmd
test

---
title: "noam_data_exploration_v1"
author: "Noam Altman-Kurosaki"
date: "2024-05-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(glmmTMB)
library(car)
library(effects)
library(DHARMa)
library(stringr)
library(MuMIn)
```

```{r }
data <- read.csv(here::here("data", "algaeonly_cleaned.csv"), stringsAsFactors = F) # replacements don't work when strings are factors

str(data)
# test
#pivot long

long_data <- pivot_longer(
  data = data,
  cols = 8:77,
  names_to = "Taxa",
  values_to = "Percent_cover")

@@

site_level <- transect_level %>%
  dplyr::group_by(Year, Site, Habitat, Taxa) %>%
  dplyr::summarise(Mean_site_cover = mean(Mean_cover), 
            Site_se = sd(Mean_cover)/sqrt(n()))
```

```{r diversity indicies, fig.height = 15, fig.width=20}
meta_cols <- colnames(data[1:7]) 
species_cols <- colnames(data[8:77])

wide_transect_level <- data %>%
  dplyr::group_by(Year, Site, Habitat, Site_habitat, Transect) %>%
  dplyr::summarize(across(all_of(species_cols), mean, na.rm = TRUE))

alpha_diversity <-
  data.frame(
    Year = as.factor(wide_transect_level$Year),
    Site = as.factor(wide_transect_level$Site),
    Habitat = as.factor(wide_transect_level$Habitat),
    Site_habitat = as.factor(wide_transect_level$Site_habitat),
    Transect = as.factor(wide_transect_level$Transect),
    Site_habitat_transect = as.factor(paste(wide_transect_level$Site_habitat, wide_transect_level$Transect)),
    Shannon = diversity(wide_transect_level[5:74], "shannon"),
    Richness = rowSums(wide_transect_level[5:74] > 0)
  )

alpha_diversity_means <- alpha_diversity %>%
  dplyr::group_by(Year, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()))

site_colours <- c("LTER 1" = "cornflowerblue",
               "LTER 2" = "darkblue",
               "LTER 3" = "yellowgreen",
               "LTER 4" = "forestgreen",
               "LTER 5" = "mediumorchid",
               "LTER 6" = "orange")
habitat_linetypes <- c("Fringing" = "solid",
                       "Backreef" = "longdash",
                       "Outer 10" = "twodash",
                       "Outer 17" = "dotted")
habitat_shapes <- c("Fringing" = 16,
                     "Backreef" = 17,
                     "Outer 10" = 1,
                     "Outer 17" = 2)

(richness_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = Year, y = Shannon_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8) +
  geom_line(aes(x = Year, y = Shannon_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Shannon_mean - Shannon_se , ymax = Shannon_mean + Shannon_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "b. Shannon\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

```{r basic alpha diversity stats}
# these violate all sorts of assumptions I'm just messing around
richness_mod <- glmmTMB(Richness ~ Year*Site*Habitat + (1|Site_habitat_transect), data = alpha_diversity, family = poisson())
Anova(richness_mod)

# shannon_mod <- glmmTMB(Shannon ~ Year*Site*Habitat + (1|Site_habitat_transect), data = alpha_diversity, family = Gamma(link = "log")) # 0 values fuck this up. Return to this when not tired. 
```

```{r nmds, fig.height = 15, fig.width = 15}
species_matrix <- as.matrix(wide_transect_level[6:74])
rownames(species_matrix) <- paste(wide_transect_level$Year, wide_transect_level$Site_habitat, wide_transect_level$Transect)
rows_to_remove <- rowSums(species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
species_matrix <- species_matrix[!rows_to_remove,] 
wide_transect_level_subset <- wide_transect_level[!rows_to_remove,] # make sure rows are also removed from dataframe
wide_transect_level_subset$Stratum <- paste(wide_transect_level_subset$Year, "-", wide_transect_level_subset$Site, "-", wide_transect_level_subset$Habitat )

# identifying species w/ > 10% cover for plotting later
# NOTE: Used ChatGPT to generate these next few lines of code but it looked good so I went with it
columns_to_keep <- apply(species_matrix, 2, function(x) any(x >= 10))
abundant_species_matrix <- as.data.frame(species_matrix)[ , c(TRUE, columns_to_keep)]
abundant_species_names <- colnames(abundant_species_matrix)

# WARNING - Before you run this, know that it will take a long time
set.seed(1032) 
# ord2 <- metaMDS(species_matrix, dist	= "bray",	trymax	= 100,	k	= 2) # looks like it automatically wisconsin transforms now?
# ord2 # stress = 0.22
# stressplot(ord2)

>>>>

---
title: "noam_data_exploration_v1"
author: "Noam Altman-Kurosaki"
date: "2024-05-09"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(glmmTMB)
library(car)
library(effects)
library(DHARMa)
library(stringr)
library(MuMIn)
library(codyn)
library(nlme)
library(mgcv)
library(emmeans)
```

```{r import and format data}
# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned.csv"), stringsAsFactors = F)

not_algae <- c("Ascidian", "Bare.Space", "Coral", "Coral.Rubble", "Corallimorpharia", "Heteractis.sp.", "Millepora.platyphylla", "No.data", "Sand", "Sacrophyton.sp.", "Shell.Debris", "Soft.Coral", "Sponge", "Tridacna.sp.", "Zooxanthellae", "row_sums")

data <- data[ , -which(names(data) %in% not_algae)]

# adding island side - adapted from Julianna's code
data <- data %>% mutate(Island_side = case_when(Site == "LTER 1" ~ "North",
                                 Site == "LTER 2" ~ "North",
                                 Site == "LTER 3" ~ "East",
                                 Site == "LTER 4" ~ "East",
                                 Site == "LTER 5" ~ "West",
                                 Site == "LTER 6" ~ "West"))
data$Island_side_habitat <- paste(data$Island_side, data$Habitat)

str(data)

meta_cols <- c(colnames(data[1:7] ), "Island_side", "Island_side_habitat")
species_cols <- colnames(data[8:77])

#pivot long 
long_data <- pivot_longer(
  data = data,
  cols = 8:77,
  names_to = "Taxa",
  values_to = "Percent_cover")

@@

# site_level <- transect_level %>%
#   dplyr::group_by(Year, Site, Habitat, Taxa) %>%
#   dplyr::summarise(Mean_site_cover = mean(Mean_cover), 
#             Site_se = sd(Mean_cover)/sqrt(n()))
```

```{r diversity indicies, fig.height = 15, fig.width=20}

wide_transect_level <- data %>%
  dplyr::group_by(Year, Island_side, Island_side_habitat, Site, Habitat, Site_habitat, Transect) %>%
  dplyr::summarize(across(all_of(species_cols), mean, na.rm = TRUE))

alpha_diversity <-
  data.frame(
    Year = as.factor(wide_transect_level$Year),
    Island_side = as.factor(wide_transect_level$Island_side),
    Island_side_habitat = as.factor(wide_transect_level$Island_side_habitat),
    Site = as.factor(wide_transect_level$Site),
    Habitat = as.factor(wide_transect_level$Habitat),
    Site_habitat = as.factor(wide_transect_level$Site_habitat),
    Transect = as.factor(wide_transect_level$Transect),
    Site_habitat_transect = as.factor(paste(wide_transect_level$Site_habitat, wide_transect_level$Transect)),
    Shannon = diversity(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)], "shannon"),
    Richness = rowSums(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)] > 0)
  )

alpha_diversity_means <- alpha_diversity %>%
  dplyr::group_by(Year, Island_side, Island_side_habitat, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()))

site_colours <- c("LTER 1" = "cornflowerblue",
               "LTER 2" = "darkblue",
               "LTER 3" = "yellowgreen",
               "LTER 4" = "forestgreen",
               "LTER 5" = "mediumorchid",
               "LTER 6" = "orange")
island_side_colours <- c("North" = "darkblue",
                         "East" = "mediumorchid",
                         "West" = "orange")
habitat_linetypes <- c("Fringing" = "solid",
                       "Backreef" = "longdash",
                       "Outer 10" = "twodash",
                       "Outer 17" = "dotted")
habitat_shapes <- c("Fringing" = 16,
                     "Backreef" = 17,
                     "Outer 10" = 1,
                     "Outer 17" = 2)

(richness_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = Year, y = Shannon_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Shannon_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Shannon_mean - Shannon_se , ymax = Shannon_mean + Shannon_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "b. Shannon\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

```{r alpha diversity quadrat level, fig.height = 15, fig.width=20}
alpha_diversity_quad <-
  data.frame(
    Year = as.numeric(data$Year),
    Location = as.factor(data$Location),
    Island_side = as.factor(data$Island_side),
    Island_side_habitat = as.factor(data$Island_side_habitat),
    Site = as.factor(data$Site),
    Habitat = as.factor(data$Habitat),
    Site_habitat = as.factor(data$Site_habitat),
    Transect = as.factor(data$Transect),
    Site_habitat_transect = as.factor(paste(data$Site_habitat, data$Transect)),
    Shannon = diversity(data[ , -which(names(data) %in% meta_cols)], "shannon"),
    Richness = rowSums(data[ , -which(names(data) %in% meta_cols)] > 0)
  )

alpha_diversity_quad_means <- alpha_diversity_quad %>%
  dplyr::group_by(Year, Island_side, Island_side_habitat, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()))

(richness_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness quad level\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = Year, y = Shannon_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Shannon_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Shannon_mean - Shannon_se , ymax = Shannon_mean + Shannon_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "b. Shannon quad level\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

```
```{r compare quad and transect level plots, fig.height = 30, fig.width= 40}
ggarrange(richness_plot + rremove("x.text"), richness_quad_plot + rremove("x.text"),
          shannon_plot,shannon_quad_plot,
          common.legend = T,
          nrow = 2, ncol = 2)

```{r gls approach}

# FRAMING: What is the general trend in algal alpha diversity over the course of the time series? Increasing, decreasing, neutral?

columns <- c("Location", "Intercept", "Slope") # names of columns for dataframe
gls_coefficients <- # new dataframe name
  data.frame(matrix(nrow = length(unique(alpha_diversity_quad$Location)), # 1200 quadrats = 1200 different time series
                    ncol = length(columns)) # specified to make sure number of columns doesn't exceed what was specified
             )
colnames(gls_coefficients) <- columns # change column names to what was specified
gls_coefficients$Location <- levels(alpha_diversity_quad$Location) # populate Location column with each quad/time series

for(i in 1:length(unique(alpha_diversity_quad$Location)) ){ # for each  quadrat
  # create a new dataframe for the quadrat of the current iteration
  newData <- subset(alpha_diversity_quad, Location == levels(alpha_diversity_quad$Location)[i])

  # fit gls for current time series
  loop_mod <- gls(Richness ~ Year, correlation =corAR1(form=~Year), data = newData)

  # populate appropriate row and column with the relevant coeffcients from gls
  gls_coefficients[i, 2] <- unname(coef(loop_mod)[1])
  gls_coefficients[i, 3] <- unname(coef(loop_mod)[2])

}

hist(gls_coefficients$Slope) # looks basically normal, bounded -Inf, Inf

hist(gls_coefficients$Intercept) # same with intercept - although it probably shouldn't be

# Start adding back in relevant metadata - probably a more elegant way to do this!!!
habitat_vector <- c(rep("Backreef", 50), rep("Fringing", 50), rep("Outer 10", 50), rep("Outer 17", 50))
transect_sequence <- c(rep(1,10),
                       rep(2,10),
                       rep(3,10),
                       rep(4,10),
                       rep(5,10))
extra_meta <- 
  data.frame(Site = as.factor(c(rep("LTER 1", 200),
                                rep("LTER 2", 200),
                                rep("LTER 3", 200),
                                rep("LTER 4", 200),
                                rep("LTER 5", 200),
                                rep("LTER 6", 200)) ),
             Island_side = as.factor(c(rep("North", 400),
                                     rep("East", 400),
                                     rep("West", 400))
                                     ),
             Habitat = as.factor(rep(habitat_vector, 6 )),
             Transect = as.factor(rep(transect_sequence, 240)),
             Location = levels(alpha_diversity_quad$Location) )

gls_coefficients_merged <- merge(gls_coefficients, extra_meta, by = "Location")

slope_mod <- glmmTMB(Slope ~ Habitat*Island_side + (1|Site/Transect), data = gls_coefficients_merged)
summary(slope_mod) # pretty low variance in random effects
plot(gls_res <- simulateResiduals(slope_mod)) # there are issues, but I can't figure out what they are...
# I think this is due to 1200 obs in model
hist(residuals(slope_mod))
plot(predict(slope_mod) ~ residuals(slope_mod))
Anova(slope_mod)
#                        Chisq Df Pr(>Chisq)    
# Island_side           16.075  2  0.0003232 ***
# Habitat             1301.260  3  < 2.2e-16 ***
# Island_side:Habitat  799.995  6  < 2.2e-16 ***
r.squaredGLMM(slope_mod) # 0.16, 0.18
plot(allEffects(slope_mod))
# Backreef: richness increases on East and North shore, but relatively constant west side
# Fringe: richness decreasing on east and west side, but relatvely constant on north shore
# Outer 10: Richness neutral/decreasing on east side, but neutral/increasing on north and west side
# Outer 17: Richness decreasing north/east side, but neutral west side
# Worth noting trends are slight - arguable whether they are biologically meaningful

slope_emm <- emmeans(slope_mod, ~ Habitat*Island_side) # create emmeans object
emmeans::contrast(slope_emm, "consec", simple = "each", combine = TRUE, adjust = "bonferroni")
# Note: not sure I want to include this
```

```{r gls plot, fig.height=20, fig.width=20}
# not sure how to visualize here... thinking something like Dornelas et al. 2014 Fig 2a
# I reached out to them about getting their code for this

ggplot(data = gls_coefficients_merged) +
  geom_histogram(aes(x = Slope), bins = 30) +
  geom_vline(xintercept = 0, size = 2, linetype = 2) +
  facet_grid(Habitat ~ Island_side) +
#  theme_classic() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain")
  )

```{r gls intercept, fig.height=20, fig.width=20}
intercept_mod <- glmmTMB(Intercept ~ Habitat*Island_side + (1|Site/Transect), data = gls_coefficients_merged)
summary(intercept_mod) # a lot of variance w/i transect, less so w/i site
plot(gls_int_res <- simulateResiduals(intercept_mod)) # same issues as above

hist(residuals(intercept_mod))
plot(predict(intercept_mod) ~ residuals(intercept_mod))
Anova(intercept_mod)
#                        Chisq Df Pr(>Chisq)    
# Habitat             1307.838  3  < 2.2e-16 ***
# Island_side           26.358  2  1.889e-06 ***
# Habitat:Island_side  789.627  6  < 2.2e-16 ***
r.squaredGLMM(intercept_mod) # 0.16, 0.18
plot(allEffects(intercept_mod))

ggplot(data = gls_coefficients_merged) +
  geom_histogram(aes(x = Intercept), bins = 30) +
  geom_vline(xintercept = 0, size = 2, linetype = 2) +
  facet_grid(Habitat ~ Island_side) +
#  theme_classic() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain")
  )

```{r gam approach}
# NOTE: Not sure where I'm going with this yet
# Idea is to look at smoother estimates for each time series to get an idea for the "shape" of the time series
# Not sure if worthwhile, might skip

# The below takes 17 years to run in and of itself
# gam_model <- gam(Richness ~ s(Year) + s(Year, by = Habitat) + 
#                  s(Island_side, bs = "re") + s(Transect, bs = "re") + s(Location, bs = "re"),
#                  data = alpha_diversity_quad, family = poisson())

gam_columns <- c("Location", "Intercept", "edf") # names of columns for dataframe
gam_coefficients <- # new dataframe name - same workflow as above
  data.frame(matrix(nrow = length(unique(alpha_diversity_quad$Location)),
                    ncol = length(gam_columns))
             )
colnames(gam_coefficients) <- gam_columns # change column names to what was specified
gam_coefficients$Location <- levels(alpha_diversity_quad$Location) # populate Location column with each quad/time series

# same workflow as above
for(i in 1:length(unique(alpha_diversity_quad$Location)) ){ # for each  quadrat
  # create a new dataframe for the quadrat of the current iteration
  newData <- subset(alpha_diversity_quad, Location == levels(alpha_diversity_quad$Location)[i])

  # fit gam for current time series
  loop_mod <- gamm(Richness ~ s(Year), data = newData, correlation = corCAR1(form = ~ Year), method = "REML", family = poisson)
  loop_sum <- summary(loop_mod$gam) # store summary data
  # populate appropriate row and column with the relevant coeffcients from gls
  gam_coefficients[i, 2] <- unname(loop_sum$p.coeff)
  gam_coefficients[i, 3] <- unname(loop_sum$edf)

}

hist(gam_coefficients$Intercept)
hist(gam_coefficients$edf) # generally around 1 except for an outlier
# LTER 5 Fringing Reef Algae Transect 3 Quad 6 - a quadrat that has almost no algae the entire time series
# issues with quadrat - smaller sample = more 0s, makes modeling trickier
hist(subset(gam_coefficients, Location != "LTER 5 Fringing Reef Algae Transect 3 Quad 6")$edf)
sum(gam_coefficients$edf > 1.01) # all but 18/1200 quadrats can be fit by a smoother with edf < 1.01
# trends more straightforward than I anticipated?
# argument for just looking at mean richness and slope?
```

```{r glmm approach}
alpha_diversity_quad$Year_factor <- as.factor(alpha_diversity_quad$Year) # for glmmTMB ar1 structure
# Framing - is adiv generally higher in certain habitats/regions?

richness_mod <- glmmTMB(
  Richness ~ Island_side*Habitat + (1 | Island_side/Site/Transect/Location) + ar1(Year_factor + 0 | Location),
  family = poisson,
  data = alpha_diversity_quad,
  control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000))
) # overspecified but going with it for now
summary(richness_mod) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod)
plot(allEffects(richness_mod))

```
```{r turnover troubleshooting, fig.height = 30, fig.width=25}

transect_level$Site_habitat <-  paste(transect_level$Site, transect_level$Habitat)
transect_level$Stratum <- paste(transect_level$Site, "-", transect_level$Habitat, "-", transect_level$Transect)

# long_data$Site_habitat <-  paste(long_data$Site, long_data$Habitat)
# long_data$Stratum <- paste(long_data$Site, "-", long_data$Habitat, "-", long_data$Transect)

# turnover analyses break when I try to do it at the quadrat level
# checking if it's because there are quadrats that have effecively 0% algae
total_cover_per_quadrat_year <- long_data %>%
  group_by(Location, Year) %>%
  summarize(total_cover = sum(Percent_cover), .groups = 'drop')

# Filter quadrats where the total percent cover is 0%
zero_cover_quadrats <- total_cover_per_quadrat_year %>%
  filter(total_cover == 0)

# Count the number of years each quadrat had 0% cover
zero_cover_count <- zero_cover_quadrats %>%
  group_by(Location) %>%
  summarize(years_with_zero_cover = n())

# Filter quadrats that had 0% cover for multiple years
quadrats_multiple_years_zero_cover <- zero_cover_count %>%
  filter(years_with_zero_cover > 1)

length(unique(quadrats_multiple_years_zero_cover$Location)) # 144/1200 quadrats have multiple years with no algal cover

# compare this to the number of transects with 0 algae cover
total_cover_per_transect_year <- transect_level %>%
  group_by(Stratum, Year) %>%
  summarize(total_cover = sum(Mean_cover), .groups = 'drop')

# Filter quadrats where the total percent cover is 0%
zero_cover_transects <- total_cover_per_transect_year %>%
  filter(total_cover == 0)

# Count the number of years each quadrat had 0% cover
zero_cover_transect_count <- zero_cover_transects %>%
  group_by(Stratum) %>%
  summarize(years_with_zero_cover = n())

# Filter quadrats that had 0% cover for multiple years
transects_multiple_years_zero_cover <- zero_cover_transect_count %>%
  filter(years_with_zero_cover > 1)

# NO transects ever have more than one year with 0 algal cover, as opposed to 144 quadrats
# Fairly certain this is why the turnover analyses are breaking

# identify quadrats missing timepoints
years <- unique(long_data$Year)
quadrats <- unique(long_data$Location)

# Create a dataframe with all combinations of quadrats and years
all_combinations <- expand.grid(Location = quadrats, Year = years)

# Identify the missing combinations by performing an anti-join with your actual data
missing_timepoints <- anti_join(all_combinations, long_data, by = c("Location", "Year"))

# Display the missing timepoints
print(missing_timepoints)

problem_children <- c("LTER 5 Fringing Reef Algae Transect 3 Quad 9", "LTER 5 Fringing Reef Algae Transect 3 Quad 8", "LTER 5 Fringing Reef Algae Transect 3 Quad 6", "LTER 5 Fringing Reef Algae Transect 3 Quad 3", "LTER 5 Fringing Reef Algae Transect 3 Quad 10")

long_data_sub <- long_data[long_data$Location != problem_children, ] # remove problem children
# still get the same error when I try to run at quadrat level
# I think it's because of the 0s
# Error in names(output) <- `*vtmp*` : 
#  'names' attribute [1] must be the same length as the vector [0]
```

```{r turnover}
# the turnover function calculates turnover as the proportion or species that either appear or disappear between timepoints
total_turnover<-
  turnover(
  df = long_data_sub,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Percent_cover",
  replicate.var = "Location",
  metric = "total"
)

appearance <-
  turnover(
  df = long_data,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Percent_cover",
  replicate.var = "Location",
  metric = "appearance"
)

disappearance <-
  turnover(
  df = long_data,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Mean_cover",
  replicate.var = "Stratum",
  metric = "disappearance"
)
